---
title: "Bornholm GC4 Factor Page"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Read This Page In 30 Seconds

This page explains one factor from four angles:

1. where it is high/low on the map,
2. how unusual values are (`z-score`),
3. how it differs by cluster (`class_km`),
4. which indicators define it (loadings across `k`).

Rule of thumb:
high score = "more of this factor pattern", low score = "less".

```{r}
suppressPackageStartupMessages({
  library(DBI)
  library(RPostgres)
  library(sf)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(leaflet)
  library(scales)
  library(knitr)
})
```

```{r}
indicator_key <- function(x) {
  case_when(
    str_detect(x, "roads") ~ "roads",
    str_detect(x, "fastboende|fastboendebefolkning") ~ "fastboende",
    str_detect(x, "ecology|oekologiskfor") ~ "ecology",
    str_detect(x, "cultural|historical|kultur") ~ "cultural",
    TRUE ~ "other"
  )
}

indicator_palette <- c(
  roads = "#1f78b4",
  fastboende = "#e31a1c",
  ecology = "#33a02c",
  cultural = "#ff7f00",
  other = "#6a3d9a"
)

class_palette <- c(
  "0" = "#8dd3c7",
  "1" = "#fdb462",
  "2" = "#bebada",
  "3" = "#fccde5",
  "4" = "#b3de69",
  "5" = "#ffed6f",
  "6" = "#bc80bd",
  "7" = "#80b1d3",
  "8" = "#fb8072",
  "9" = "#ccebc5"
)

factor_id <- 'F4'
valid_factors <- c("F1", "F2", "F3", "F4", "F5")
if (!factor_id %in% valid_factors) stop("Invalid factor param")

# Locate DB helper from current location
root_candidates <- c(
  normalizePath(getwd(), winslash = "/", mustWork = FALSE),
  normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE),
  normalizePath(file.path(getwd(), "..", ".."), winslash = "/", mustWork = FALSE)
)
db_connect_candidates <- unlist(lapply(root_candidates, function(r) {
  c(
    file.path(r, "R/db_connect.R"),
    file.path(r, "databas/generell_databas_setup/R/db_connect.R"),
    file.path(r, "generell_databas_setup/R/db_connect.R")
  )
}))
db_connect_path <- db_connect_candidates[file.exists(db_connect_candidates)][1]
if (is.na(db_connect_path) || !nzchar(db_connect_path)) stop("Could not find db_connect.R")
source(db_connect_path)
project_root <- dirname(dirname(db_connect_path))

env_candidates <- c(
  Sys.getenv("PIPELINE_ENV_PATH", unset = ""),
  file.path(project_root, ".env"),
  file.path(project_root, "..", ".env")
)
env_candidates <- env_candidates[nzchar(env_candidates)]
cfg_env <- env_candidates[file.exists(env_candidates)][1]
if (is.na(cfg_env) || !nzchar(cfg_env)) stop("No .env file found for DB connection")

schema <- Sys.getenv("PIPELINE_SCHEMA", "h3")
view_gc4 <- Sys.getenv("QGIS_VIEW_FACTORS_GC4", "v_bornholm_r8_factors_gc4")

con <- if ("path" %in% names(formals(connect_pg))) {
  connect_pg(cfg_env)
} else if ("cfg_env" %in% names(formals(connect_pg))) {
  connect_pg(cfg_env)
} else {
  if (exists("load_pg_env", mode = "function")) load_pg_env(cfg_env)
  DBI::dbConnect(
    RPostgres::Postgres(),
    dbname = Sys.getenv("PGDATABASE"),
    host = Sys.getenv("PGHOST", "localhost"),
    port = as.integer(Sys.getenv("PGPORT", "5432")),
    user = Sys.getenv("PGUSER"),
    password = Sys.getenv("PGPASSWORD")
  )
}

gc4 <- st_read(con, query = sprintf("SELECT * FROM %s.%s", schema, view_gc4), quiet = TRUE) |>
  st_transform(4326)

loadings_candidates <- c(
  file.path(project_root, "jyp_note_book_geocontext", "bornholm_r8_factor_loadings_gc4.csv"),
  file.path(project_root, "bornholm_r8_factor_loadings_gc4.csv")
)
loadings_path <- loadings_candidates[file.exists(loadings_candidates)][1]
if (is.na(loadings_path) || !nzchar(loadings_path)) stop("Could not find bornholm_r8_factor_loadings_gc4.csv")
loadings_df <- read.csv(loadings_path, stringsAsFactors = FALSE)

build_factor_sf <- function(fid) {
  gc4 |>
    mutate(
      factor_value = .data[[fid]],
      factor_z = as.numeric(scale(factor_value)),
      z_band = case_when(
        factor_z >= 1.5 ~ "hotspot (z >= 1.5)",
        factor_z <= -1.5 ~ "coldspot (z <= -1.5)",
        TRUE ~ "mid"
      )
    )
}
```

## Spatial View: `r factor_id`

Use layer toggles to switch between:

- raw factor score,
- standardized score (`z-score`),
- cluster overlay.

Hotspots/coldspots are simple flags of unusually high/low values.

```{r}
factor_id <- 'F4'
factor_sf <- build_factor_sf(factor_id)
max_abs <- max(abs(factor_sf$factor_value), na.rm = TRUE)
if (!is.finite(max_abs) || max_abs <= 0) max_abs <- 1

pal_score <- colorNumeric(palette = "RdBu", domain = c(-max_abs, max_abs), reverse = TRUE)
pal_z <- colorNumeric(palette = "RdBu", domain = c(-3, 3), reverse = TRUE)

m <- leaflet(factor_sf) |>
  addProviderTiles("CartoDB.Positron", group = "CartoDB") |>
  addProviderTiles("OpenStreetMap.Mapnik", group = "OSM") |>
  addPolygons(
    group = "Factor score",
    fillColor = ~pal_score(factor_value),
    fillOpacity = 0.7,
    color = "#333333",
    weight = 0.35,
    popup = ~paste0("hex: ", hex_id, "<br>", factor_id, ": ", round(factor_value, 3), "<br>z: ", round(factor_z, 3))
  ) |>
  addPolygons(
    group = "Factor z-score",
    fillColor = ~pal_z(pmax(pmin(factor_z, 3), -3)),
    fillOpacity = 0.7,
    color = "#333333",
    weight = 0.35,
    popup = ~paste0("hex: ", hex_id, "<br>", factor_id, ": ", round(factor_value, 3), "<br>z: ", round(factor_z, 3))
  )

if ("class_km" %in% names(factor_sf)) {
  class_vals <- sort(unique(as.character(factor_sf$class_km)))
  pal_class <- colorFactor(unname(class_palette[class_vals]), domain = class_vals)
  m <- m |>
    addPolygons(
      group = "Cluster class_km",
      fillColor = ~pal_class(class_km),
      fillOpacity = 0.45,
      color = "#333333",
      weight = 0.35,
      popup = ~paste0("hex: ", hex_id, "<br>class_km: ", class_km, "<br>", factor_id, ": ", round(factor_value, 3))
    ) |>
    addLegend("bottomleft", pal = pal_class, values = ~class_km, title = "class_km", group = "Cluster class_km")
}

hot_sf <- factor_sf |> filter(z_band == "hotspot (z >= 1.5)")
cold_sf <- factor_sf |> filter(z_band == "coldspot (z <= -1.5)")
if (nrow(hot_sf) > 0) {
  m <- m |>
    addPolygons(data = hot_sf, group = "Hotspots", fillColor = "#B2182B", fillOpacity = 0.65, color = "#7F0000", weight = 0.4)
}
if (nrow(cold_sf) > 0) {
  m <- m |>
    addPolygons(data = cold_sf, group = "Coldspots", fillColor = "#2166AC", fillOpacity = 0.65, color = "#053061", weight = 0.4)
}

m |>
  addLegend("bottomright", pal = pal_score, values = ~factor_value, title = paste0(factor_id, " score"), group = "Factor score") |>
  addLegend("topright", pal = pal_z, values = ~pmax(pmin(factor_z, 3), -3), title = paste0(factor_id, " z-score"), group = "Factor z-score") |>
  addLayersControl(
    baseGroups = c("CartoDB", "OSM"),
    overlayGroups = c("Factor score", "Factor z-score", "Cluster class_km", "Hotspots", "Coldspots"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

```{r}
hot_n <- nrow(factor_sf |> filter(z_band == "hotspot (z >= 1.5)"))
cold_n <- nrow(factor_sf |> filter(z_band == "coldspot (z <= -1.5)"))
cat(paste0(
  "**Key takeaway:** `", factor_id, "` has ", hot_n, " hotspot hexagons and ",
  cold_n, " coldspot hexagons using z-threshold \u00b11.5."
))
```

## Distribution and Class Context

The histogram shows the overall spread of `r factor_id` across Bornholm.  
The boxplot shows how that spread differs between clusters.

```{r}
factor_id <- 'F4'
factor_sf <- build_factor_sf(factor_id)
hist_df <- factor_sf |>
  st_drop_geometry() |>
  mutate(factor_selected = factor_value) |>
  select(hex_id, class_km, factor_selected, factor_z)

ggplot(hist_df, aes(x = factor_selected)) +
  geom_histogram(bins = 35, fill = "#4C78A8", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  labs(title = paste0(factor_id, " score distribution"), x = factor_id, y = "Hex count") +
  theme_minimal(base_size = 12)
```

```{r}
stats <- hist_df |>
  summarise(
    med = median(factor_selected, na.rm = TRUE),
    q1 = quantile(factor_selected, 0.25, na.rm = TRUE),
    q3 = quantile(factor_selected, 0.75, na.rm = TRUE)
  )
cat(paste0(
  "**Key takeaway:** Typical `", factor_id, "` value is ",
  round(stats$med, 3), " (middle 50% from ", round(stats$q1, 3),
  " to ", round(stats$q3, 3), ")."
))
```

```{r}
factor_id <- 'F4'
if ("class_km" %in% names(hist_df)) {
  ggplot(hist_df, aes(x = factor(class_km), y = factor_selected, fill = factor(class_km))) +
    geom_boxplot(show.legend = FALSE, outlier.alpha = 0.2) +
    scale_fill_manual(values = class_palette, drop = FALSE) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
    labs(title = paste0(factor_id, " by class_km"), x = "class_km", y = factor_id) +
    theme_minimal(base_size = 12)
}
```

```{r}
if ("class_km" %in% names(hist_df)) {
  class_rank <- hist_df |>
    group_by(class_km) |>
    summarise(med = median(factor_selected, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(med))
  cat(paste0(
    "**Key takeaway:** Cluster ", class_rank$class_km[1],
    " has the highest median `", factor_id, "` (",
    round(class_rank$med[1], 3), ")."
  ))
}
```

## `r factor_id` Loadings vs K (Mean)

This plot shows whether interpretation of `r factor_id` stays similar when
neighborhood size changes (`k = 10 ... 1000`).  
More stable lines = more robust factor meaning.

```{r}
factor_id <- 'F4'
plot_df <- loadings_df |>
  filter(str_detect(variable, "^mean_")) |>
  mutate(
    k = as.integer(str_extract(variable, "(?<=_k)[0-9]+$")),
    series = str_remove(variable, "^mean_"),
    series = str_remove(series, "_k[0-9]+$"),
    series_key = indicator_key(series)
  ) |>
  filter(!is.na(k)) |>
  select(variable, series, series_key, k, all_of(factor_id)) |>
  rename(loading = all_of(factor_id))

end_labels <- plot_df |>
  group_by(series) |>
  filter(k == max(k)) |>
  slice(1) |>
  ungroup()

ggplot(plot_df, aes(x = k, y = loading, color = series_key, group = series)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "black") +
  geom_line(linewidth = 1.0) +
  geom_point(size = 1.8) +
  geom_text(data = end_labels, aes(label = series), hjust = -0.08, size = 3.2, show.legend = FALSE) +
  scale_x_continuous(breaks = sort(unique(plot_df$k))) +
  scale_color_manual(values = indicator_palette, drop = FALSE) +
  coord_cartesian(ylim = c(-1, 1), clip = "off") +
  labs(
    title = paste0("Factor loadings vs K neighbors (mean): ", factor_id),
    x = "K neighbors",
    y = "Factor loading",
    color = "Series"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 120, 5.5, 5.5)
  )
```

```{r}
stab <- plot_df |>
  group_by(series) |>
  summarise(delta = max(loading, na.rm = TRUE) - min(loading, na.rm = TRUE), .groups = "drop") |>
  arrange(delta)
cat(paste0(
  "**Key takeaway:** Most stable contributor across k is `",
  stab$series[1], "`."
))
```

## Top Contributing Variables

These tables are the practical "why" behind the factor:

- top positive loadings = indicators that increase the factor,
- top negative loadings = indicators that decrease the factor.

```{r}
factor_id <- 'F4'
contrib_df <- loadings_df |>
  transmute(
    variable,
    loading = .data[[factor_id]],
    abs_loading = abs(loading),
    type = case_when(
      str_detect(variable, "^mean_") ~ "mean",
      str_detect(variable, "^std_") ~ "std",
      TRUE ~ "other"
    ),
    k = as.integer(str_extract(variable, "(?<=_k)[0-9]+$"))
  ) |>
  arrange(desc(abs_loading))

cat("Top positive loadings\n")
kable(contrib_df |> arrange(desc(loading)) |> select(variable, loading, type, k) |> head(12), digits = 3)

cat("\nTop negative loadings\n")
kable(contrib_df |> arrange(loading) |> select(variable, loading, type, k) |> head(12), digits = 3)
```

```{r}
top_pos <- contrib_df |> arrange(desc(loading)) |> slice(1)
top_neg <- contrib_df |> arrange(loading) |> slice(1)
cat(paste0(
  "**Key takeaway:** Strongest positive driver is `", top_pos$variable,
  "` (", round(top_pos$loading, 3), "), while strongest negative is `",
  top_neg$variable, "` (", round(top_neg$loading, 3), ")."
))
```

```{r}
if (exists("con") && DBI::dbIsValid(con)) DBI::dbDisconnect(con)
```
