---
title: "Bornholm GC4 Snapshot"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

## Read This First

This report translates many technical steps into a simple question:
which places on Bornholm are similar, different, and why?

How to read:

1. `Cluster` = groups of hexagons with similar factor profiles (`F1..F5`).
2. `Factor` = a data pattern that combines many indicators into one score.
3. High/low factor scores do not mean "good/bad" by themselves; they mean
"more/less of this pattern."

## What Is A Cluster Built On, And Why?

Clusters (`class_km`) are built from the factor scores (`F1..F5`), not from one
single raw layer. This is useful because:

- it reduces many variables to a smaller set of patterns,
- it groups areas that behave similarly across several indicators,
- it makes map interpretation easier than comparing dozens of raw layers.

Use clusters as a navigation aid, not as a final truth.

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(DBI)
  library(RPostgres)
  library(sf)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(leaflet)
  library(scales)
})

indicator_key <- function(x) {
  case_when(
    str_detect(x, "roads") ~ "roads",
    str_detect(x, "fastboende|fastboendebefolkning") ~ "fastboende",
    str_detect(x, "ecology|oekologiskfor") ~ "ecology",
    str_detect(x, "cultural|historical|kultur") ~ "cultural",
    TRUE ~ "other"
  )
}

indicator_palette <- c(
  roads = "#1f78b4",
  fastboende = "#e31a1c",
  ecology = "#33a02c",
  cultural = "#ff7f00",
  other = "#6a3d9a"
)

class_palette <- c(
  "0" = "#8dd3c7",
  "1" = "#fdb462",
  "2" = "#bebada",
  "3" = "#fccde5",
  "4" = "#b3de69",
  "5" = "#ffed6f",
  "6" = "#bc80bd",
  "7" = "#80b1d3",
  "8" = "#fb8072",
  "9" = "#ccebc5"
)
```

```{r}
#| label: connect
# Locate and source shared DB helper (walk up from current dir)
root_candidates <- c(
  normalizePath(getwd(), winslash = "/", mustWork = FALSE),
  normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE),
  normalizePath(file.path(getwd(), "..", ".."), winslash = "/", mustWork = FALSE)
)
db_connect_candidates <- unlist(lapply(root_candidates, function(r) {
  c(
    file.path(r, "R/db_connect.R"),
    file.path(r, "databas/generell_databas_setup/R/db_connect.R"),
    file.path(r, "generell_databas_setup/R/db_connect.R")
  )
}))
db_connect_path <- db_connect_candidates[file.exists(db_connect_candidates)][1]
if (is.na(db_connect_path) || !nzchar(db_connect_path)) {
  stop("Could not find db_connect.R")
}
source(db_connect_path)
project_root <- dirname(dirname(db_connect_path))

env_candidates <- c(
  Sys.getenv("PIPELINE_ENV_PATH", unset = ""),
  file.path(project_root, ".env"),
  file.path(project_root, "..", ".env")
)
env_candidates <- env_candidates[nzchar(env_candidates)]
cfg_env <- env_candidates[file.exists(env_candidates)][1]
if (is.na(cfg_env) || !nzchar(cfg_env)) {
  stop("No .env file found for DB connection")
}
schema <- Sys.getenv("PIPELINE_SCHEMA", "h3")
view_gc4 <- Sys.getenv("QGIS_VIEW_FACTORS_GC4", "v_bornholm_r8_factors_gc4")
features_table <- Sys.getenv("GC_FEATURES_TABLE", "bornholm_r8_geocontext_features")
factors_table <- Sys.getenv("GC4_FACTORS_TABLE", "bornholm_r8_factors_gc4")

con <- if ("path" %in% names(formals(connect_pg))) {
  connect_pg(cfg_env)
} else if ("cfg_env" %in% names(formals(connect_pg))) {
  connect_pg(cfg_env)
} else {
  if (exists("load_pg_env", mode = "function")) load_pg_env(cfg_env)
  DBI::dbConnect(
    RPostgres::Postgres(),
    dbname = Sys.getenv("PGDATABASE"),
    host = Sys.getenv("PGHOST", "localhost"),
    port = as.integer(Sys.getenv("PGPORT", "5432")),
    user = Sys.getenv("PGUSER"),
    password = Sys.getenv("PGPASSWORD")
  )
}
```

```{r}
#| label: load-data
qry <- sprintf("SELECT * FROM %s.%s", schema, view_gc4)
gc4 <- st_read(con, query = qry, quiet = TRUE)
gc4 <- st_transform(gc4, 4326)

cat("Rows:", nrow(gc4), "\n")
cat("Columns:", ncol(gc4), "\n")
```

```{r}
#| label: load-feature-factor-join
feat <- NULL
feat_source <- "none"

# 1) Try DB table join first
qry_join <- sprintf(
  "SELECT f.*, s.\"F1\", s.\"F2\", s.\"F3\", s.\"F4\", s.\"F5\", s.class_km
   FROM %s.%s f
   LEFT JOIN %s.%s s ON f.hex_id = s.hex_id",
  schema, features_table, schema, factors_table
)
feat_db <- tryCatch(dbGetQuery(con, qry_join), error = function(e) NULL)

has_k_cols <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(FALSE)
  any(str_detect(names(df), "^mean_gc_.*_k[0-9]+$"))
}

if (has_k_cols(feat_db)) {
  feat <- feat_db
  feat_source <- "database"
} else {
  # 2) Fallback to notebook CSV outputs
  ctx_candidates <- c(
    file.path(project_root, "jyp_note_book_geocontext", "bornholm_points_with_context_gc4.csv"),
    file.path(project_root, "bornholm_points_with_context_gc4.csv")
  )
  score_candidates <- c(
    file.path(project_root, "jyp_note_book_geocontext", "bornholm_r8_factor_scores_gc4.csv"),
    file.path(project_root, "bornholm_r8_factor_scores_gc4.csv")
  )
  ctx_path <- ctx_candidates[file.exists(ctx_candidates)][1]
  score_path <- score_candidates[file.exists(score_candidates)][1]

  if (!is.na(ctx_path) && nzchar(ctx_path)) {
    feat_ctx <- read.csv(ctx_path, stringsAsFactors = FALSE)
    if (!is.na(score_path) && nzchar(score_path)) {
      feat_scores <- read.csv(score_path, stringsAsFactors = FALSE)
      feat <- feat_ctx |>
        left_join(feat_scores, by = "hex_id")
      feat_source <- "csv_context+scores"
    } else {
      feat <- feat_ctx
      feat_source <- "csv_context"
    }
  } else if (!is.null(feat_db)) {
    feat <- feat_db
    feat_source <- "database_no_k_cols"
  }
}

if (is.null(feat)) stop("Could not load feature data from DB or CSV fallback")
cat("Feature source:", feat_source, "\n")
cat("Feature rows:", nrow(feat), "\n")
cat("Feature columns:", ncol(feat), "\n")
```

## Cluster Distribution (`class_km`)

This bar chart shows how many hexagons belong to each cluster.  
If one cluster is much larger than others, it likely captures the "typical"
landscape profile.

```{r}
if ("class_km" %in% names(gc4)) {
  gc4 |>
    st_drop_geometry() |>
    count(class_km, name = "n_hex") |>
    ggplot(aes(x = factor(class_km), y = n_hex, fill = factor(class_km))) +
    geom_col(show.legend = FALSE) +
    scale_fill_manual(values = class_palette, drop = FALSE) +
    scale_y_continuous(labels = comma) +
    labs(x = "class_km", y = "Hex count") +
    theme_minimal(base_size = 12)
} else {
  cat("class_km column not found.\n")
}
```

```{r}
if ("class_km" %in% names(gc4)) {
  cls <- gc4 |>
    st_drop_geometry() |>
    count(class_km, name = "n_hex") |>
    mutate(share = n_hex / sum(n_hex)) |>
    arrange(desc(n_hex))
  top <- cls[1, ]
  cat(paste0(
    "**Key takeaway:** Largest group is cluster `", top$class_km,
    "` with ", scales::percent(top$share, accuracy = 0.1),
    " of all hexagons."
  ))
}
```

## Cluster Map (`class_km`)

This map shows where each cluster appears geographically.  
Look for contiguous zones (coherent areas) versus scattered patches (local pockets).

```{r}
if ("class_km" %in% names(gc4)) {
  class_vals <- sort(unique(as.character(gc4$class_km)))
  pal <- colorFactor(unname(class_palette[class_vals]), domain = class_vals)

  leaflet(gc4) |>
    addProviderTiles("CartoDB.Positron") |>
    addPolygons(
      fillColor = ~pal(class_km),
      fillOpacity = 0.65,
      color = "#444444",
      weight = 0.4,
      popup = ~paste0("hex: ", hex_id, "<br>class_km: ", class_km)
    ) |>
    addLegend("bottomright", pal = pal, values = ~class_km, title = "class_km")
} else {
  cat("class_km column not found.\n")
}
```

```{r}
if ("class_km" %in% names(gc4)) {
  ncls <- gc4 |>
    st_drop_geometry() |>
    summarise(n = n_distinct(class_km)) |>
    pull(n)
  cat(paste0(
    "**Key takeaway:** The map contains ", ncls,
    " cluster types. Use this layer first to locate broad spatial patterns."
  ))
}
```

## Factor Summary

These summaries show the spread of each factor score across all hexagons
(minimum, median, maximum).  
A wider spread means the factor separates places more strongly.

```{r}
fcols <- grep("^F[0-9]+$", names(gc4), value = TRUE)
if (length(fcols) > 0) {
  gc4 |>
    st_drop_geometry() |>
    select(all_of(fcols)) |>
    summary()
} else {
  cat("No factor columns (F1..Fn) found in the view.\n")
}
```

```{r}
if (length(fcols) > 0) {
  spread_df <- gc4 |>
    st_drop_geometry() |>
    summarise(across(all_of(fcols), ~sd(.x, na.rm = TRUE))) |>
    pivot_longer(everything(), names_to = "factor", values_to = "sd_val") |>
    arrange(desc(sd_val))
  cat(paste0(
    "**Key takeaway:** `", spread_df$factor[1],
    "` has the widest spread (SD=", round(spread_df$sd_val[1], 2),
    "), so it separates places most strongly."
  ))
}
```

## Bornholm Factor Loadings Matrix

This table shows which original indicators contribute most to each factor.  
Large positive/negative values mean stronger influence on that factor.

```{r}
#| label: loadings-matrix
loadings_candidates <- c(
  file.path(project_root, "jyp_note_book_geocontext", "bornholm_r8_factor_loadings_gc4.csv"),
  file.path(project_root, "bornholm_r8_factor_loadings_gc4.csv"),
  "jyp_note_book_geocontext/bornholm_r8_factor_loadings_gc4.csv",
  "bornholm_r8_factor_loadings_gc4.csv"
)
loadings_path <- loadings_candidates[file.exists(loadings_candidates)][1]

if (is.na(loadings_path) || !nzchar(loadings_path)) {
  cat("Could not find bornholm_r8_factor_loadings_gc4.csv\n")
} else {
  loadings_df <- read.csv(loadings_path, stringsAsFactors = FALSE)
  cat("Loadings file:", loadings_path, "\n")
  cat("Rows:", nrow(loadings_df), "Cols:", ncol(loadings_df), "\n")
  knitr::kable(loadings_df, digits = 3)
}
```

```{r}
#| label: loadings-heatmap
if (exists("loadings_df")) {
  fcols_load <- intersect(c("F1", "F2", "F3", "F4", "F5"), names(loadings_df))
  loadings_long <- loadings_df |>
    select(variable, all_of(fcols_load)) |>
    pivot_longer(cols = all_of(fcols_load), names_to = "factor", values_to = "loading")

  ggplot(loadings_long, aes(x = factor, y = variable, fill = loading)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    labs(x = "Factor", y = "Variable", fill = "Loading") +
    theme_minimal(base_size = 10) +
    theme(axis.text.y = element_text(size = 7))
}
```

The heatmap is the same information as the matrix, but visual:
strong colors mean stronger loadings.

```{r}
if (exists("loadings_df")) {
  fcols_load <- intersect(c("F1", "F2", "F3", "F4", "F5"), names(loadings_df))
  top_by_factor <- lapply(fcols_load, function(f) {
    row <- loadings_df |>
      transmute(variable, loading = .data[[f]], abs_loading = abs(loading)) |>
      arrange(desc(abs_loading)) |>
      slice(1)
    paste0(f, " \u2192 ", row$variable, " (", round(row$loading, 3), ")")
  })
  cat(paste0("**Key takeaway:** Strongest single loading per factor: ", paste(unlist(top_by_factor), collapse = "; "), "."))
}
```

## Factor Loadings vs K Neighbors (Mean)

This plot shows whether factor meaning stays stable when neighborhood size (`k`)
changes.  
Stable lines across `k` mean more robust interpretation.

```{r}
#| label: loadings-vs-k-mean
if (exists("loadings_df")) {
  fcols_load <- intersect(c("F1", "F2", "F3", "F4", "F5"), names(loadings_df))

  loadings_mean_long <- loadings_df |>
    filter(str_detect(variable, "^mean_")) |>
    mutate(
      k = as.integer(str_extract(variable, "(?<=_k)[0-9]+$")),
      series = str_remove(variable, "^mean_"),
      series = str_remove(series, "_k[0-9]+$"),
      indicator = indicator_key(series)
    ) |>
    pivot_longer(cols = all_of(fcols_load), names_to = "factor", values_to = "loading") |>
    filter(!is.na(k))

  if (nrow(loadings_mean_long) > 0) {
    end_labels <- loadings_mean_long |>
      group_by(factor, series) |>
      filter(k == max(k)) |>
      slice(1) |>
      ungroup()

    ggplot(loadings_mean_long, aes(x = k, y = loading, color = indicator, group = series)) +
      geom_hline(yintercept = 0, linewidth = 0.35, color = "black") +
      geom_line(linewidth = 0.8) +
      geom_point(size = 1.2) +
      geom_text(
        data = end_labels,
        aes(label = series),
        hjust = -0.08,
        size = 2.9,
        show.legend = FALSE
      ) +
      facet_wrap(~ factor, ncol = 1) +
      scale_color_manual(values = indicator_palette, drop = FALSE) +
      scale_x_continuous(breaks = sort(unique(loadings_mean_long$k))) +
      coord_cartesian(ylim = c(-1, 1), clip = "off") +
      labs(
        title = "Factor loadings vs K neighbors (mean)",
        x = "K neighbors",
        y = "Factor loadings",
        color = "Series"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        legend.position = "none",
        plot.margin = margin(5.5, 90, 5.5, 5.5)
      )
  } else {
    cat("No mean_* loadings available for K-plot.\n")
  }
}
```

```{r}
if (exists("loadings_mean_long") && nrow(loadings_mean_long) > 0) {
  stab <- loadings_mean_long |>
    group_by(factor, series) |>
    summarise(range_abs = max(loading, na.rm = TRUE) - min(loading, na.rm = TRUE), .groups = "drop") |>
    group_by(factor) |>
    summarise(mean_range = mean(range_abs, na.rm = TRUE), .groups = "drop") |>
    arrange(mean_range)
  cat(paste0(
    "**Key takeaway:** Most stable factor across k is `", stab$factor[1],
    "` (smallest average loading drift)."
  ))
}
```

## Per-Factor Pages

- Interpretation hub: [bornholm_gc4_index.html](./bornholm_gc4_index.html)
- [F1 page](./bornholm_gc4_F1.html)
- [F2 page](./bornholm_gc4_F2.html)
- [F3 page](./bornholm_gc4_F3.html)
- [F4 page](./bornholm_gc4_F4.html)
- [F5 page](./bornholm_gc4_F5.html)

## Mean Indicator Trends Across K

These plots show how average indicator values change when context size (`k`) grows.  
Use them to see if local patterns persist at broader neighborhood scales.

```{r}
#| label: prep-k-trends
k_target <- c(10, 50, 100, 250, 1000)

mean_cols <- names(feat)[str_detect(names(feat), "^mean_gc_.*_k[0-9]+$")]
if (length(mean_cols) == 0) {
  cat("No mean_gc_*_k* columns found in feature table.\n")
} else {
  mean_long <- feat |>
    select(any_of(c("hex_id", "class_km")), all_of(mean_cols)) |>
    pivot_longer(
      cols = all_of(mean_cols),
      names_to = "colname",
      values_to = "value"
    ) |>
    mutate(
      k = as.integer(str_extract(colname, "(?<=_k)[0-9]+$")),
      variable = str_remove(colname, "^mean_"),
      variable = str_remove(variable, "_k[0-9]+$"),
      indicator = indicator_key(variable)
    ) |>
    filter(k %in% k_target)
}
```

```{r}
#| label: plot-k-trends-overall
if (exists("mean_long")) {
  mean_long |>
    group_by(variable, indicator, k) |>
    summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = k, y = mean_value, color = indicator)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.6) +
    scale_color_manual(values = indicator_palette, drop = FALSE) +
    scale_x_continuous(breaks = c(10, 50, 100, 250, 1000)) +
    labs(
      x = "k (nearest neighbours)",
      y = "Mean indicator value",
      color = "Indicator"
    ) +
    theme_minimal(base_size = 12)
}
```

```{r}
if (exists("mean_long") && nrow(mean_long) > 0) {
  trend <- mean_long |>
    group_by(variable, k) |>
    summarise(v = mean(value, na.rm = TRUE), .groups = "drop") |>
    group_by(variable) |>
    summarise(delta = v[k == max(k)] - v[k == min(k)], .groups = "drop") |>
    arrange(desc(abs(delta)))
  cat(paste0(
    "**Key takeaway:** Biggest k-scale change appears in `", trend$variable[1],
    "` (\u0394=", round(trend$delta[1], 3), ")."
  ))
}
```

```{r}
#| label: plot-k-trends-by-cluster
if (exists("mean_long") && "class_km" %in% names(mean_long)) {
  mean_long |>
    filter(!is.na(class_km)) |>
    group_by(class_km, variable, k) |>
    summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = k, y = mean_value, color = factor(class_km))) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    scale_color_manual(values = class_palette, drop = FALSE) +
    scale_x_continuous(breaks = c(10, 50, 100, 250, 1000)) +
    labs(
      x = "k (nearest neighbours)",
      y = "Mean indicator value",
      color = "class_km"
    ) +
    facet_wrap(~ variable, scales = "free_y", ncol = 2) +
    theme_minimal(base_size = 11)
}
```

## Factor Correlation With Indicators Across K

These panels show how strongly each indicator tracks each factor at different `k`.
They help answer: which indicators actually drive each factor, and is that stable?

```{r}
#| label: factor-k-correlation
fcols_feat <- intersect(c("F1", "F2", "F3", "F4", "F5"), names(feat))
if (length(fcols_feat) == 0 || length(mean_cols) == 0) {
  cat("Missing factors or mean indicator columns for correlation analysis.\n")
} else {
  corr_rows <- list()
  i <- 1L
  for (f in fcols_feat) {
    for (mc in mean_cols) {
      k <- as.integer(str_extract(mc, "(?<=_k)[0-9]+$"))
      if (!k %in% k_target) next
      var_name <- str_remove(mc, "^mean_")
      var_name <- str_remove(var_name, "_k[0-9]+$")
      cc <- suppressWarnings(cor(feat[[f]], feat[[mc]], use = "pairwise.complete.obs"))
      corr_rows[[i]] <- data.frame(
        factor = f,
        variable = var_name,
        indicator = indicator_key(var_name),
        k = k,
        corr = as.numeric(cc)
      )
      i <- i + 1L
    }
  }
  corr_df <- bind_rows(corr_rows)
}
```

```{r}
#| label: plot-factor-k-correlation
if (exists("corr_df") && nrow(corr_df) > 0) {
  corr_df |>
    ggplot(aes(x = k, y = corr, color = indicator)) +
    geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    scale_color_manual(values = indicator_palette, drop = FALSE) +
    scale_x_continuous(breaks = c(10, 50, 100, 250, 1000)) +
    labs(
      x = "k (nearest neighbours)",
      y = "Correlation with factor score",
      color = "Indicator"
    ) +
    facet_wrap(~ factor, ncol = 2) +
    theme_minimal(base_size = 11)
}
```

```{r}
if (exists("corr_df") && nrow(corr_df) > 0) {
  top_corr <- corr_df |>
    mutate(abs_corr = abs(corr)) |>
    arrange(desc(abs_corr)) |>
    slice(1)
  cat(paste0(
    "**Key takeaway:** Strongest factor-indicator link is ",
    top_corr$factor, " with `", top_corr$variable,
    "` at k=", top_corr$k, " (r=", round(top_corr$corr, 3), ")."
  ))
}
```

```{r}
#| label: disconnect
if (exists("con") && DBI::dbIsValid(con)) {
  DBI::dbDisconnect(con)
}
```
